% ************************** Thesis Acknowledgements *****************************
\begin{acknowledgements}

This project is my most substantial piece of work and I've learned an incredible
amount from this all too short period. I am greatly looking forward to continuing
development of the software packages that this project created and hope to pick
up the analysis where this project leaves off over summer.

I owe a massive thank you to both my dissertation supervisor,
Amanda Clare at Aberystwyth and my contact at the Sanger Institute, Joshua
Randall for their excellent guidance, suggestions and general advice which
helped me steer the project through the unknown.

Additionally I'd like to express my gratitude to the Wellcome Trust Sanger Institute for
supporting the project and granting me access to their data sets and
computational resources without which the project would not have been possible.

I would like to thank my best friend and ladyfiend for keeping me sane and
putting up with me for the duration of the project, despite irrationally holding her
personally responsible for the fundamental quirks of biology. Thank you Louise, I
wouldn't have finished this without you.

Thanks also to Emily and Dan of Cat House for delivering my daily dose of
caffeine and for joining in on pizza evenings, at least we were able to enjoy the
stress in good company.

To my fellow constituents of \#\#zuzakistan, thanks for the occasional
distribution of tea, muffins and typewriters and the distracting discussions
on the wonders of space travel.


Finally thank you to my family and friends for all your support and confidence
in me, throughout my degree. It's been an excellent experience and I look forward
to continuing my adventures at Aberystwyth.

\end{acknowledgements}
